### 2019-08-07 submission/manuscript.tex
Changes in response to Pat's comments & edits

\section{\texorpdfstring{Proton pump inhibitor administration does not
promote \emph{Clostridium difficile} colonization in a murine
model}{Proton pump inhibitor administration does not promote Clostridium difficile colonization in a murine model}}\label{proton-pump-inhibitor-administration-does-not-promote-clostridium-difficile-colonization-in-a-murine-model}

@@

Proton pump inhibitor (PPI) use has been associated with microbiota
alterations and susceptibility to \emph{Clostridium difficile}
infections (CDIs) in humans. We assessed how PPI treatment alters the
fecal microbiota and whether PPIs promote CDIs in a mouse model. Mice
receiving the PPI treatment were gavaged with 40 mg/kg of omeprazole
during a 7-day pretreatment phase, the day of \emph{C. difficile}
challenge, and the following 9 days. We found that mice treated with
omeprazole were not colonized by \emph{C. difficile}. When omeprazole
treatment was combined with a single clindamycin treatment, one cage of
mice remained resistant to \emph{C. difficile} colonization, while the
other cage was colonized. Treatming mice with clindamycin and no
omeprazole followed by challenge resulted in \emph{C. difficile}
colonization. 16S rRNA gene sequencing analysis revealed that omeprazole
had minimal impact on the structure of the murine microbiota throughout
the 16 days of PPI exposure. These results suggest PPI treatment alone
is not sufficient to disrupt microbiota resistance to \emph{C. difficle}
infection in mice that are normally resistant in the absence of
antibiotic treatment.

@@

Antibiotics are a major risk factor for \emph{Clostridium difficile}
infections (CDIs), but other factors may also contribute. In
epidemiological studies, PPI use has been associated with CDI incidence
and recurrence. Proton pump inhibitors (PPIs) have also been associated
with alterations in the human intestinal microbiota in observational and
interventional studies. We evaluated the effects of the PPI omeprazole
on the structure of the murine intestinal microbiota and its ability to
disrupt colonization resistance to \emph{C. difficile}. We found the PPI
treatment had minimal impact on the murine fecal microbiota and did not
promote \emph{C. difficile} colonization. Further studies are needed to
determine whether other factors such as the composition of the starting
bacterial community, comorbidities, and use of additional medications
contribute to the association between PPIs and CDIs seen in humans or
whether aspects of murine physiology may limit its utility to test these
types of hypotheses.

@@

Antibiotics have a large impact on the intestinal microbiome and are a
primary risk factor for developing \emph{Clostridium difficile}
infections (CDIs) (1). It is less clear whether other human medications
that impact the microbiota also influence \emph{C. difficile}
colonization resistance. Multiple epidemiological studies have suggested
an association between proton pump inhibitor (PPI) use and incidence or
recurrence of CDIs (2--5). There have also been a number of large cohort
studies as well as interventional clinical trials that have demonstrated
that specific alterations in the intestinal microbiome were associated
with PPI use (4, 6). PPI-associated microbiota changes have been
attributed to the ability of PPIs to increase stomach acid pH which may
promote the survival of oral and pathogenic bacteria (4, 6). Human fecal
microbiota changes with PPI use include increases in Enterococcaceae,
Lactobacillaceae, Micrococcaceae, Staphylococcaceae and Streptococcaceae
and decreases in Ruminococcaceae (6, 7). Several of these taxa have been
associated with C. difficile colonization in humans (8)\\
Unfortunately, most of the studies suggesting a link between PPIs and
\emph{C. difficile} were retrospective and did not evaluate the
microbiome (2, 3, 5). Thus, it is unclear whether the gastrointestinal
microbiome changes associated with PPI use explain the association
between PPIs and CDIs. Additionally, epidemiological studies have a
limited capacity to address potential confounders and comorbidities in
patients that were on PPIs and developed CDIs or recurrent CDIs (2, 5).
Here, we evaluated the impact of a daily high dose PPI treatment on the
murine microbiome and susceptibility to \emph{C. difficile} colonization
in relation to clindamycin, an antibiotic that perturbs the microbiome
enough to allow \emph{C. difficile} to colonize but is mild enough that
\emph{C. difficile} is cleared within 10 days (9).

\textbf{Murine fecal microbiomes were minimally affected by PPI
treatment } To test whether PPI treatment alters the microbiome and
promotes susceptibility to CDIs, we gavaged mice with 40 mg/kg of
omeprazole for 7 days before \emph{C. difficile} challenge (Figure 1A).
A principle coordinates analysis (PCoA) of the Bray-Curtis distances
over the initial 7 days of treatment revealed the bacterial communities
of PPI-treated mice remained relatively unchanged relative to animals
that did not recieve the PPI (Figure 1B). We observed no significant
fluctuations in the relative abundance of those taxa previously shown to
respond to PPI treatment throughout the course of the 16-day experiment
(Figure 1C, Figure S1A-B). We also observed no significant fluctuations
in relative abundances at the family and genus level over the beginning,
day of \emph{C. difficile} challenge, and end of the experiment for the
PPI-treated mice (Figure 1D, Figure S1C). Next, the fecal bacterial
communities of PPI-treated mice were compared to a group of mice that
received a single 10 mg/kg dose of clindamycin and a group that received
both PPI and clindamycin. In contrast to the dynamics of mice only
receiving omeprazole, the microbiomes from the 2 groups of mice that
received clindamycin started to shift away from the PPI-treated mice and
the samples from previous days one day after treatment (Figure 1D).
These results demonstrated that the PPI treatment alone had a minimal
impact on the murine fecal bacterial community relative to the impact of
clindamycin treatment.

\textbf{PPI-treatment did not promote susceptibility to \emph{C.
difficile infection} in mice} Next, we examined whether PPI treatment
altered susceptibility to \emph{C. difficile} infection in mice. After
PPI treatment or clindamycin treatment, mice were challenged with
\emph{C. difficile} 630 spores. While \emph{C. difficile} colonized the
clindamycin treated mice, it was unable to colonize the PPI-treated mice
(Figure 2A). Interestingly, only 1 cage of mice that received both the
PPI and clindamycin were colonized, while the mice in other cage wwere
resistant (Figure 2A). The greatest shifts in bacterial communities
occurred in the clindamycin-treated mice (Figure 2B, Figure S2).
Regardless of whether the mice became colonized, all of the mice had
cleared \emph{C. difficile} within 5 days (Figure 2A), suggesting that
omeprazole did not affect the rate of clearance. Our results suggest
that PPI treatment had no effects on bacterial community resistance to
\emph{C. difficile} colonization in mice. Instead most of the
differences between our 3 treatment groups appear to be driven by
clindamycin administration (Figure 2C-D) and included previously
described decreases in \emph{Alistipes}, \emph{Barnesiella},
Porhyromonadaceae, Ruminococcaceae (1). These findings demonstrated that
high dose PPI treatment did not promote susceptibility to \emph{C.
difficile} colonization.

@@

not promote \emph{C. difficile} infection in mice. Our findings that PPI
treatment had minimal impact on the fecal microbiome were comparable to
another PPI mouse study that indicated PPIs had more of an effect on the
small intestinal microbiota compared to the fecal microbiota (10). The
same group demonstrated PPI treatment increased the stomach pH in mice
(10), which may improve survival of bacteria passing through the
stomach. We did not find significant changes for the taxa observed to be
significantly impacted by PPI use in human studies. However, 3 of the
human-associated taxa were absent or at low abundance in our mice.
Additionally, some of the significance of PPI associations in human
interventional trials appears to be driven by a handful of specific taxa
with overall differences on a PCoA plot difficult to distinguish (11).
The PPI-treated mice fecal communities did not change overall and no
differences in the relative abundances of individual families or genera
were observed over time. While our fecal microbiota findings are
comparable to what's been shown in another mouse study, whether
PPI-induced changes in specific bacterial abundances observed in humans
play a role in CDIs remains to be determined.\\
Although there have been a few \emph{C. difficile} mouse model studies
that have demonstrated PPIs have some effect on CDIs with or without
additional antibiotic treatment (13--15), there were some key
methodological differences between these studies and our own. One group
administered 0.5 mg/kg of the PPI lansoprazole daily for 2 weeks to mice
and then challenged with \emph{C. difficile} demonstrated that PPI
treatment alone resulted in detectable \emph{C. difficile} in the stool
1 week after challenge, but also showed there was detectable \emph{C.
difficile} in mice not treated with antibiotics (13, 14). The presence
of \emph{C. difficile} in mice not treated with antibiotics may be
partly attributed to the higher dose of vegetative cells
(10\textsuperscript{8} CFU) used, but nevertheless makes it difficult to
interpret the extent to which PPI treatment contributed to \emph{C.
difficile} colonization in their mouse model (13, 14). In contrast, we
have previously shown mice from our colony that were not given
antibiotics were resistant to \emph{C. difficile} 630 when challenged
with 10\textsuperscript{3} spores (16) and previous work by our
collaborators demonstrated mice not given antibiotics were resistant to
10\textsuperscript{5} vegetative cells of \emph{C. difficile } VPI 10463
(17). The other PPI CDI mouse study also used a higher dose of
10\textsuperscript{6} spores or 10\textsuperscript{7} vegetative cells
to challenge antibiotic-treated mice or mice treated with both
antibiotics and the PPI esomeprazole (40mg/kg dose) for 2 days and
demonstrated the antibiotic/PPI-treated mice developed more severe CDIs
(15). We tested the same high 40 mg/kg PPI dose and expanded
pre-treatment to 7 days before challenge to test the impact of PPI
treatment alone on our CDI mouse model.\\
Our study extended the previous work examining PPIs and \emph{C.
difficile} in mice by evaluating the contribution of the intestinal

@@

experiment and in contrast to previous work, PPIs did not alter \emph{C.
difficile} colonization resistance. 16S rRNA sequencing suggested that
\emph{Streptococcus} and \emph{Enterococcus} are rare genera in our
C57BL/6 mouse colony and variation in these 2 genera have been observed
across other facilities and vendors (18, 19). These genera could be
important contributors to the associations between PPIs and CDIs in
humans, and could be a contributing factor to our observation that PPI
treatment had no effect on \emph{C. difficile} colonization in our CDI
mouse model.\\

@@

\textbf{Animals} All mouse experiments were performed with 7- to
12-week-old C57BL/6 male and female mice. Each experimental group of
mice was spit between 2 cages with 2-3 mice housed per cage and male and
female mice housed separately. All animal experiments were approved by
the University of Michigan Animal Care and Use Committee (IACUC) under
protocol number PRO00006983.

\textbf{Drug treatments} Omeprazole (Sigma Aldrich) was prepared in a
vehicle solution of 40\% polyethylene glycol 400 (Sigma-Aldrich) in
phosphate buffered saline. Omeprazole was prepared from 20 mg/mL frozen
aliquots and diluted to an 8mg/mL prior to gavage. All mice received 40
mg/kg omeprazole or vehicle solution once per day through the duration
of the experiment with treatment starting 7 days before \emph{C.
difficile} challenge (Figure 1A). One day prior to \emph{C. difficile}
challenge, 2 groups of mice received an intraperitoneal injection of 10
mg/kg clindamycin or sterile saline vehicle. All drugs were filter
sterilized through a 0.22 micron syringe filter before administration to
animals.

\textbf{\emph{C. difficile} infection model} Mice were challenged with
\emph{C. difficile} 630 seven days after the start of omeprazole
treatment and one day after clindamycin treatment. Mice were challenged
with 10\textsuperscript{3} spores in ultrapure distilled water. Stool
samples were collected for 16S rRNA sequencing or \emph{C. difficile}
CFU quantification throughout the duration of the experiments at the
indicated timepoints (Figure 1A). Samples for 16S rRNA sequencing were
flash frozen in liquid nitrogen and stored at -80°C until DNA
extraction, while samples for CFU quantification were transferred into
an anaerobic chamber and serially diluted in PBS. Diluted samples were
plated on TCCFA (taurocholate, cycloserine, cefoxitin, fructose agar)
plates and incubated at 37°C for 24 hours under anaerobic conditions to
quantify \emph{C. difficile} CFU.

\textbf{16S rRNA gene sequencing} DNA for 16S rRNA gene sequencing was
extracted from 10-50 mg fecal pellet from each mouse using the DNeasy
Powersoil HTP 96 Kit (Qiagen) and an EpMotion 5075 automated pipetting
system (Eppendorf). The ZymoBIOMICS\textsuperscript{TM} Microbial
Community DNA Standard (Zymo, CA, USA) was used as a mock community (21)
and water was used as a negative control. The V4 hypervariable region of
the 16S rRNA gene was amplified with Accuprime Pfx DNA polymerase
(Thermo Fisher Scientific) using previously described custom barcoded
primers (22). The 16S rRNA amplicon library was sequenced with the MiSeq
(Illumina). Amplicons were cleaned up and normalized with the SequalPrep
Normalization Plate Kit (ThermoFisher Scientific) and pooled amplicons
were quantified with the KAPA library quantification kit (KAPA
Biosystems).

\textbf{16S rRNA gene sequence analysis}Mothur (v1.40.5) was used for
all sequence processing steps (23) using the previously published
protocol (22). In brief, forward and reverse reads for each sample were
combined and low-quality sequences and chimeras were removed. Duplicate
sequences were merged, before taxonomy assignment using a modified
version (v16) of the Ribosomal Database Project reference database
(v11.5) with an 80\% cutoff. Operational taxonomic units (OTUs) were
assigned with the opticlust clustering algorithm using a 97\% similarity
threshold. To adjust for uneven sequencing across samples, all samples
were rarefied to 3000 sequences, 1,000 times. PCoAs were generated based
on Bray-Curtis distance. R (v.3.5.1) was used to generate figures and
perform statistical analysis.

\textbf{Statistical Analysis} To test for differences in relative
abundances in families and genera across our 3 different treatment
groups (Clindamycin, Clindamycin + PPI, and PPI) or within the PPI
treatment group across 3 timepoints (Day -7, 0, and 9), we used a
Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple
comparisons.

\textbf{Code availability} The code for all sequence processing and
analysis step as well as an Rmarkdown version of this manuscript is
available at
\url{https://github.com/SchlossLab/Tomkovich_PPI_XXXX_2019}.

\textbf{Data availability} The 16S rRNA sequencing data have been
deposited in the NCBI Sequence Read Archive (Accession no. PRJNA554866).

@@

\hypertarget{ref-maier2018extensive}{}
7. \textbf{Maier L}, \textbf{Pruteanu M}, \textbf{Kuhn M},
\textbf{Zeller G}, \textbf{Telzerow A}, \textbf{Anderson EE},
\textbf{Brochado AR}, \textbf{Fernandez KC}, \textbf{Dose H},
\textbf{Mori H}, \textbf{others}. 2018. Extensive impact of
non-antibiotic drugs on human gut bacteria. Nature \textbf{555}:623.

\hypertarget{ref-Schuberte01021-14}{}
8. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
\textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
\textbf{Aronoff DM}, \textbf{Schloss PD}. 2014. Microbiome data
distinguish patients with clostridium difficile infection and non-c.
difficile-associated diarrhea from healthy controls. mBio \textbf{5}.
doi:\href{https://doi.org/10.1128/mBio.01021-14}{10.1128/mBio.01021-14}.

\hypertarget{ref-Jenior2018}{}
9. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
\textbf{Schloss PD}. 2018. Clostridium difficile alters the structure
and metabolism of distinct cecal microbiomes during initial infection to
promote sustained colonization. mSphere \textbf{3}.
doi:\href{https://doi.org/10.1128/msphere.00261-18}{10.1128/msphere.00261-18}.

\hypertarget{ref-Yasutomi2018}{}
10. \textbf{Yasutomi E}, \textbf{Hoshi N}, \textbf{Adachi S},
\textbf{Otsuka T}, \textbf{Kong L}, \textbf{Ku Y}, \textbf{Yamairi H},
\textbf{Inoue J}, \textbf{Ishida T}, \textbf{Watanabe D}, \textbf{Ooi
M}, \textbf{Yoshida M}, \textbf{Tsukimi T}, \textbf{Fukuda S},
\textbf{Azuma T}. 2018. Proton pump inhibitors increase the
susceptibility of mice to oral infection with enteropathogenic bacteria.
Digestive Diseases and Sciences \textbf{63}:881--889.
doi:\href{https://doi.org/10.1007/s10620-017-4905-3}{10.1007/s10620-017-4905-3}.

\hypertarget{ref-Freedberg2015}{}
11. \textbf{Freedberg DE}, \textbf{Toussaint NC}, \textbf{Chen SP},
\textbf{Ratner AJ}, \textbf{Whittier S}, \textbf{Wang TC}, \textbf{Wang
HH}, \textbf{Abrams JA}. 2015. Proton pump inhibitors alter specific
taxa in the human gastrointestinal microbiome: A crossover trial.
Gastroenterology \textbf{149}:883--885.e9.
doi:\href{https://doi.org/10.1053/j.gastro.2015.06.043}{10.1053/j.gastro.2015.06.043}.

\hypertarget{ref-Imhann2015}{}
12. \textbf{Imhann F}, \textbf{Bonder MJ}, \textbf{Vila AV}, \textbf{Fu
J}, \textbf{Mujagic Z}, \textbf{Vork L}, \textbf{Tigchelaar EF},
\textbf{Jankipersadsing SA}, \textbf{Cenit MC}, \textbf{Harmsen HJM},
\textbf{Dijkstra G}, \textbf{Franke L}, \textbf{Xavier RJ},
\textbf{Jonkers D}, \textbf{Wijmenga C}, \textbf{Weersma RK},
\textbf{Zhernakova A}. 2015. Proton pump inhibitors affect the gut
microbiome. Gut \textbf{65}:740--748.
doi:\href{https://doi.org/10.1136/gutjnl-2015-310376}{10.1136/gutjnl-2015-310376}.

@@

\hypertarget{ref-Sze2019}{}
21. \textbf{Sze MA}, \textbf{Schloss PD}. 2019. The impact of DNA
polymerase and number of rounds of amplification in PCR on 16S rRNA gene
sequence data. mSphere \textbf{4}.
doi:\href{https://doi.org/10.1128/msphere.00163-19}{10.1128/msphere.00163-19}.

>>>>

\section{\texorpdfstring{The proton pump inhibitor omeprazole does not
promote \emph{Clostridium difficile} colonization in a murine
model}{The proton pump inhibitor omeprazole does not promote Clostridium difficile colonization in a murine model}}\label{the-proton-pump-inhibitor-omeprazole-does-not-promote-clostridium-difficile-colonization-in-a-murine-model}

@@

Proton pump inhibitor (PPI) use has been associated with microbiota
alterations and susceptibility to \emph{Clostridium difficile}
infections (CDIs) in humans. We assessed how PPI treatment alters the
fecal microbiota and whether treatment promotes CDIs in a mouse model.
Mice receiving a PPI treatment were gavaged with 40 mg/kg of omeprazole
during a 7-day pretreatment phase, the day of \emph{C. difficile}
challenge, and the following 9 days. We found that mice treated with
omeprazole were not colonized by \emph{C. difficile}. When omeprazole
treatment was combined with a single clindamycin treatment, one cage of
mice remained resistant to \emph{C. difficile} colonization, while the
other cage was colonized. Treating mice with only clindamycin followed
by challenge resulted in \emph{C. difficile} colonization. 16S rRNA gene
sequencing analysis revealed that omeprazole had minimal impact on the
structure of the murine microbiota throughout the 16 days of omeprazole
exposure. These results suggest omeprazole treatment alone is not
sufficient to disrupt microbiota resistance to \emph{C. difficle}
infection in mice that are normally resistant in the absence of
antibiotic treatment.

@@

Antibiotics are the primary risk factor for \emph{Clostridium difficile}
infections (CDIs), but other factors may also increase a person's risk.
In epidemiological studies, PPI use has been associated with CDI
incidence and recurrence. Proton pump inhibitors (PPIs) have also been
associated with alterations in the human intestinal microbiota in
observational and interventional studies. We evaluated the effects of
the PPI omeprazole on the structure of the murine intestinal microbiota
and its ability to disrupt colonization resistance to \emph{C.
difficile}. We found omeprazole treatment had minimal impact on the
murine fecal microbiota and did not promote \emph{C. difficile}
colonization. Further studies are needed to determine whether other
factors such as the composition of the starting bacterial community,
comorbidities, and use of additional medications contribute to the
association between PPIs and CDIs seen in humans or whether aspects of
murine physiology may limit its utility to test these types of
hypotheses.

@@

Antibiotics have a large impact on the intestinal microbiome and are a
primary risk factor for developing \emph{Clostridium difficile}
infections (CDIs) (1). It is less clear whether other human medications
that impact the microbiota also influence \emph{C. difficile}
colonization resistance. Multiple epidemiological studies have suggested
an association between proton pump inhibitor (PPI) use and incidence or
recurrence of CDIs (2--5). There have also been a number of large cohort
studies and interventional clinical trials that demonstrated specific
alterations in the intestinal microbiome were associated with PPI use
(4, 6). PPI-associated microbiota changes have been attributed to the
ability of PPIs to increase stomach acid pH which may promote the
survival of oral and pathogenic bacteria (4, 6). In human fecal samples,
PPI use results in increases in Enterococcaceae, Lactobacillaceae,
Micrococcaceae, Staphylococcaceae and Streptococcaceae and decreases in
Ruminococcaceae (6--9). Several of these taxa have also been associated
with \emph{C. difficile} colonization in humans (10).\\
Unfortunately, most of the studies suggesting a link between PPIs and
\emph{C. difficile} were retrospective and did not evaluate changes in
the microbiome (2, 3, 5). Thus, it is unclear whether the
gastrointestinal microbiome changes associated with PPI use explain the
association between PPIs and CDIs. Additionally, epidemiological studies
have a limited capacity to address potential confounders and
comorbidities in patients that were on PPIs and developed CDIs or
recurrent CDIs (2, 5). Here, we evaluated the impact of daily PPI
treatment with omeprazole on the murine microbiome and susceptibility to
\emph{C. difficile} colonization in relation to clindamycin, an
antibiotic that perturbs the microbiome enough to allow \emph{C.
difficile} to colonize but is mild enough that \emph{C. difficile} is
cleared within 10 days (11).

\textbf{Murine fecal microbiomes were minimally affected by omeprazole
treatment } To test whether omeprazole treatment alters the microbiome
and promotes susceptibility to CDIs, we gavaged mice with 40 mg/kg of
omeprazole for 7 days before \emph{C. difficile} challenge (Figure 1A).
A principle coordinates analysis (PCoA) of the Bray-Curtis distances
over the initial 7 days of treatment revealed the bacterial communities
of omeprazole-treated mice remained relatively unchanged (Figure 1B). We
observed no significant fluctuations in the relative abundance of those
taxa previously shown to respond to PPI treatment throughout the course
of the 16-day experiment (Figure 1C-D, S1). We also observed no
significant changes in relative abundances at the family and genus level
over the course of the experiment for the omeprazole-treated mice (data
not shown). These results demonstrated that the omeprazole treatment
alone had a minimal impact on the murine fecal bacterial community after
7 days of pretreatment.

\textbf{Omeprazole treatment did not promote susceptibility to \emph{C.
difficile infection} in mice} Next, we examined whether omeprazole
treatment altered susceptibility to \emph{C. difficile} infection in
mice. After omeprazole treatment or clindamycin treatment, mice were
challenged with 10\textsuperscript{3} \emph{C. difficile} 630 spores.
While \emph{C. difficile} colonized the clindamycin-treated mice, it did
not colonize the omeprazole-treated mice (Figure 2A). Interestingly,
only 1 cage of mice that received both omeprazole and clindamycin were
colonized, while the other cage of mice were resistant (Figure 2A). The
greatest shifts in bacterial communities occurred in the
clindamycin-treated mice (Figure 2B, S2). Regardless of whether the mice
became colonized, all of the mice had cleared \emph{C. difficile} within
5 days (Figure 2A), suggesting that omeprazole did not affect the rate
of clearance. Our results suggest that omeprazole treatment had no
effect on bacterial community resistance to \emph{C. difficile}
colonization in mice. Instead most of the differences between the 3
treatment groups appeared to be driven by clindamycin administration
(Figure 2C, S2). These findings demonstrated that high dose omeprazole
treatment did not promote susceptibility to \emph{C. difficile}
colonization.

@@

not promote \emph{C. difficile} infection in mice. Our findings that
omeprazole treatment had minimal impact on the fecal microbiome were
comparable to another PPI mouse study that indicated the PPI
lansoprazole had more of an effect on the small intestinal microbiota
compared to the fecal microbiota (12). The same group demonstrated
lansoprazole treatment increased the stomach pH in mice (12), which may
improve survival of bacteria passing through the stomach. We did not
find significant changes for the taxa observed to be significantly
impacted by PPI use in human studies. However, 3 of the human-associated
taxa were absent or at low abundance in our mice. Although our fecal
microbiota findings are comparable to what has been shown in another
mouse study (12), whether PPI-induced changes in specific bacterial
abundances observed in humans play a role in CDIs remains to be
determined.\\
While several \emph{C. difficile} mouse model studies have shown that
PPIs have an effect on CDIs with or without additional antibiotic
treatment (13--15), there were insufficient controls to attribute the
effect solely to PPI treatment. One group administered 0.5 mg/kg of the
PPI lansoprazole daily for 2 weeks to mice and then challenged with
\emph{C. difficile} demonstrated that PPI treatment alone resulted in
detectable \emph{C. difficile} in the stool 1 week after challenge,
however there was detectable \emph{C. difficile} in mice not treated
with antibiotics (13, 14). The other mouse study demonstrated the
antibiotic/PPI (esomeprazole)-treated mice developed more severe CDIs
compared to antibiotic-treated mice, but did not have a group treated
with just esomeprazole for comparison (15). We tested the same high 40
mg/kg PPI dose and expanded pre-treatment to 7 days before challenge to
test the impact of omeprazole treatment alone on our CDI mouse model.
Additionally, we've previously demonstrated that mice from our colony
were resistant to \emph{C. difficile} 630 colonization without
antibiotic treatment (16), ensuring there was not already partial
susceptiblity to \emph{C. difficile} before treatment. The additional
controls in our study allowed us to assess the contibution of omeprazole
alone to \emph{C. difficile} susceptibility in mice.\\
Our study also extended previous work examining PPIs and \emph{C.
difficile} in mice by incorporating the contribution of the intestinal

@@

experiment. In contrast to previous work, omeprazole did not alter
\emph{C. difficile} colonization resistance. 16S rRNA sequencing
suggested that \emph{Streptococcus} and \emph{Enterococcus} are rare
genera in our C57BL/6 mouse colony and variation in these 2 genera have
been observed across other facilities and vendors (17, 18). These genera
could be important contributors to the associations between PPIs and
CDIs in humans, and could be a contributing factor to our observation
that PPI treatment had no effect on \emph{C. difficile} colonization in
our CDI mouse model. Gastrointestinal physiological differences,
particularly the higher stomach pH in mice (pH 3-4) compared to humans
(ph 1) (19) could also explain why omeprazole had a limited impact on
the murine microbiome. The microbiota and physiological differences
between humans and mice may limit the usefulness of employing mouse
models to study the impact of PPIs on the microbiota and CDIs.\\

@@

\textbf{Animals.} All mouse experiments were performed with 7- to
12-week-old C57BL/6 male and female mice. Each experimental group of
mice was spit between 2 cages with 2-3 mice housed per cage and male and
female mice housed separately. All animal experiments were approved by
the University of Michigan Animal Care and Use Committee (IACUC) under
protocol number PRO00006983.

\textbf{Drug treatments.} Omeprazole (Sigma Aldrich) was prepared in a
vehicle solution of 40\% polyethylene glycol 400 (Sigma-Aldrich) in
phosphate buffered saline. Omeprazole was prepared from 20 mg/mL frozen
aliquots and diluted to an 8 mg/mL prior to gavage. All mice received 40
mg/kg omeprazole (a dose previously used in mouse experiments (15)) or
vehicle solution once per day through the duration of the experiment
with treatment starting 7 days before \emph{C. difficile} challenge
(Figure 1A). One day prior to \emph{C. difficile} challenge, 2 groups of
mice received an intraperitoneal injection of 10 mg/kg clindamycin or
sterile saline vehicle (11). All drugs were filter sterilized through a
0.22 micron syringe filter before administration to animals.

\textbf{\emph{C. difficile} infection model.} Mice were challenged with
\emph{C. difficile} 630 seven days after the start of omeprazole
treatment and one day after clindamycin treatment. Mice were challenged
with 10\textsuperscript{3} spores in ultrapure distilled water as
described previously (11). Stool samples were collected for 16S rRNA
sequencing or \emph{C. difficile} CFU quantification throughout the
duration of the experiments at the indicated timepoints (Figure 1A).
Samples for 16S rRNA sequencing were flash frozen in liquid nitrogen and
stored at -80°C until DNA extraction, while samples for CFU
quantification were transferred into an anaerobic chamber and serially
diluted in PBS. Diluted samples were plated on TCCFA (taurocholate,
cycloserine, cefoxitin, fructose agar) plates and incubated at 37°C for
24 hours under anaerobic conditions to quantify \emph{C. difficile} CFU.

\textbf{16S rRNA gene sequencing.} DNA for 16S rRNA gene sequencing was
extracted from 10-50 mg fecal pellet from each mouse using the DNeasy
Powersoil HTP 96 Kit (Qiagen) and an EpMotion 5075 automated pipetting
system (Eppendorf). The 16S rRNA sequencing libary was prepared as
described previously (21). In brief, the ZymoBIOMICS\textsuperscript{TM}
Microbial Community DNA Standard (Zymo, CA, USA) was used as a mock
community (22) and water was used as a negative control. The V4
hypervariable region of the 16S rRNA gene was amplified with Accuprime
Pfx DNA polymerase (Thermo Fisher Scientific) using previously described
custom barcoded primers (21). The 16S rRNA amplicon library was
sequenced with the MiSeq (Illumina). Amplicons were cleaned up and
normalized with the SequalPrep Normalization Plate Kit (ThermoFisher
Scientific) and pooled amplicons were quantified with the KAPA library
quantification kit (KAPA Biosystems).

\textbf{16S rRNA gene sequence analysis.} mothur (v1.40.5) was used for
all sequence processing steps (23) using a previously published protocol
(21). In brief, forward and reverse reads for each sample were combined
and low-quality sequences and chimeras were removed. Duplicate sequences
were merged, before taxonomy assignment using a modified version (v16)
of the Ribosomal Database Project reference database (v11.5) with an
80\% cutoff. Operational taxonomic units (OTUs) were assigned with the
opticlust clustering algorithm using a 97\% similarity threshold. To
adjust for uneven sequencing across samples, all samples were rarefied
to 3,000 sequences, 1,000 times. PCoAs were generated based on
Bray-Curtis distance. R (v.3.5.1) was used to generate figures and
perform statistical analysis.

\textbf{Statistical Analysis.} To test for differences in relative
abundances in families and genera across our 3 different treatment
groups (Clindamycin, Clindamycin + PPI, and PPI) or within the PPI
treatment group across 3 timepoints (Day -7, 0, and 9), we used a
Kruskal-Wallis test with a Benjamini-Hochberg correction for multiple
comparisons.

\textbf{Code availability.} The code for all sequence processing and
analysis step as well as an Rmarkdown version of this manuscript is
available at
\url{https://github.com/SchlossLab/Tomkovich_PPI_mSphere_2019}.

\textbf{Data availability.} The 16S rRNA sequencing data have been
deposited in the NCBI Sequence Read Archive (Accession no. PRJNA554866).

@@

\hypertarget{ref-Imhann2015}{}
7. \textbf{Imhann F}, \textbf{Bonder MJ}, \textbf{Vila AV}, \textbf{Fu
J}, \textbf{Mujagic Z}, \textbf{Vork L}, \textbf{Tigchelaar EF},
\textbf{Jankipersadsing SA}, \textbf{Cenit MC}, \textbf{Harmsen HJM},
\textbf{Dijkstra G}, \textbf{Franke L}, \textbf{Xavier RJ},
\textbf{Jonkers D}, \textbf{Wijmenga C}, \textbf{Weersma RK},
\textbf{Zhernakova A}. 2015. Proton pump inhibitors affect the gut
microbiome. Gut \textbf{65}:740--748.
doi:\href{https://doi.org/10.1136/gutjnl-2015-310376}{10.1136/gutjnl-2015-310376}.

\hypertarget{ref-Freedberg2015}{}
8. \textbf{Freedberg DE}, \textbf{Toussaint NC}, \textbf{Chen SP},
\textbf{Ratner AJ}, \textbf{Whittier S}, \textbf{Wang TC}, \textbf{Wang
HH}, \textbf{Abrams JA}. 2015. Proton pump inhibitors alter specific
taxa in the human gastrointestinal microbiome: A crossover trial.
Gastroenterology \textbf{149}:883--885.e9.
doi:\href{https://doi.org/10.1053/j.gastro.2015.06.043}{10.1053/j.gastro.2015.06.043}.

\hypertarget{ref-maier2018extensive}{}
9. \textbf{Maier L}, \textbf{Pruteanu M}, \textbf{Kuhn M},
\textbf{Zeller G}, \textbf{Telzerow A}, \textbf{Anderson EE},
\textbf{Brochado AR}, \textbf{Fernandez KC}, \textbf{Dose H},
\textbf{Mori H}, \textbf{others}. 2018. Extensive impact of
non-antibiotic drugs on human gut bacteria. Nature \textbf{555}:623.

\hypertarget{ref-Schuberte01021-14}{}
10. \textbf{Schubert AM}, \textbf{Rogers MAM}, \textbf{Ring C},
\textbf{Mogle J}, \textbf{Petrosino JP}, \textbf{Young VB},
\textbf{Aronoff DM}, \textbf{Schloss PD}. 2014. Microbiome data
distinguish patients with clostridium difficile infection and non-c.
difficile-associated diarrhea from healthy controls. mBio \textbf{5}.
doi:\href{https://doi.org/10.1128/mBio.01021-14}{10.1128/mBio.01021-14}.

\hypertarget{ref-Jenior2018}{}
11. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
\textbf{Schloss PD}. 2018. Clostridium difficile alters the structure
and metabolism of distinct cecal microbiomes during initial infection to
promote sustained colonization. mSphere \textbf{3}.
doi:\href{https://doi.org/10.1128/msphere.00261-18}{10.1128/msphere.00261-18}.

\hypertarget{ref-Yasutomi2018}{}
12. \textbf{Yasutomi E}, \textbf{Hoshi N}, \textbf{Adachi S},
\textbf{Otsuka T}, \textbf{Kong L}, \textbf{Ku Y}, \textbf{Yamairi H},
\textbf{Inoue J}, \textbf{Ishida T}, \textbf{Watanabe D}, \textbf{Ooi
M}, \textbf{Yoshida M}, \textbf{Tsukimi T}, \textbf{Fukuda S},
\textbf{Azuma T}. 2018. Proton pump inhibitors increase the
susceptibility of mice to oral infection with enteropathogenic bacteria.
Digestive Diseases and Sciences \textbf{63}:881--889.
doi:\href{https://doi.org/10.1007/s10620-017-4905-3}{10.1007/s10620-017-4905-3}.

@@

\hypertarget{ref-Kozich2013}{}
21. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq illumina sequencing platform.
Applied and Environmental Microbiology \textbf{79}:5112--5120.
doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.

\hypertarget{ref-Sze2019}{}
22. \textbf{Sze MA}, \textbf{Schloss PD}. 2019. The impact of DNA
polymerase and number of rounds of amplification in PCR on 16S rRNA gene
sequence data. mSphere \textbf{4}.
doi:\href{https://doi.org/10.1128/msphere.00163-19}{10.1128/msphere.00163-19}.
